---
title: "Indexes for simultaneous selection"
output: rmarkdown::html_vignette
link-citations: true
bibliography: metanref.bib 
vignette: >
  %\VignetteIndexEntry{Indexes for simultaneous selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



# Getting started
In this section we will use the data examples `data_ge` and `data_ge2`  provided in the metan package. For more information, please, see `?data_ge` and `?data_ge2`. Other data sets can be used provided that the following columns are in the dataset: environment, genotype, block/replicate and response variable(s).

# FAI-BLUP index
The FAI-BLUP is a multi-trait index based on factor analysis and ideotype-design recentely proposed by @Rocha2018. It is based on factor analysis, when the factorial scores of each ideotype are designed according to the desirable and undesirable factors. Then, a spatial probability is estimated based on genotype-ideotype distance, enabling genotype ranking [@Rocha2018]. The BLUPs from genotypes are obtained with the function `WAASB()` and then the function `FAI_BLUP()` is used to compute the FAI-BLUP index.

```{r, fig.height = 5, fig.width = 5.5, fig.align = "center", message=F}
library(metan)
library(magrittr) # used for the forward-pipe operator %>%

FAI = data_ge2 %>%
      WAASB(ENV, GEN, REP, c(KW, NKE, PH, EH)) %>%
      FAI_BLUP(DI = c("max", "max", "max", "min"),
               UI = c("min", "min", "min", "max"),
               SI = 15)
plot(FAI)
    ```



# Multi-trait stability index

## Based on stability only

In the following example, the selection of stable genotypes will consider the 
```{r, fig.height = 5, fig.width = 5.5, fig.align = "center" }
MTSI_index = data_ge2 %>%
             WAASB(ENV, GEN, REP,
                   resp = c(KW, NKE, PH, EH, TKW)) %>%
             MTSI()
plot(MTSI_index)
    ```


## Based on mean performance and stability

The following code considers that higher values for KW, NKE, TKW are better, and lower values for PH and EH are better. The simultaneous selection for mean performance and stability will prioritize the mean performance (mean of the variables) rather than their stability across environments. 

```{r, fig.height = 5, fig.width = 5.5, fig.align = "center" }
MTSI_index2 = data_ge2 %>%
             WAASB(ENV, GEN, REP,
                   resp = c(KW, NKE, PH, EH, TKW),
                   mresp = c(100, 100, 0, 0, 100), #Default
                   wresp = rep(65, 5)) %>% # Default is 50
             MTSI(index = "WAASBY")
plot(MTSI_index2)
    ```

# References

