---
title: "Utilities for data manipulation"
author: "Tiago Olivoto"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
link-citations: true
bibliography: metanref.bib 
vignette: >
  %\VignetteIndexEntry{Utilities for data manipulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r global_options, include = FALSE}
knitr::opts_chunk$set(comment = "")
```


# Utilities for rows and columns
## Add columns and rows
The functions `add_cols()` and `add_rows()` can be used to add columns and rows, respectively to a data frame.

```{r message=FALSE, warning=FALSE}
library(metan)
add_columns(data_ge,
            ROW_ID = 1:420)
add_columns(data_ge,
            ROW_ID = 1:420,
            before = "GEN")
```

## Select or remove columns and rows
The functions `select_cols()` and `select_rows()` can be used to select columns and rows, respectively from a data frame.

```{r}
select_cols(data_ge2, ENV, GEN) %>% head(3)
select_cols(data_ge2, ENV:GEN) %>% head(3)
select_cols(data_ge2, 1:2) %>% head(3)
select_rows(data_ge2, 1, 2, 3)
select_rows(data_ge2, 1:3)
```

To remove columns or rows, use `remove_cols()` and `remove_rows()`.
```{r}
remove_cols(data_ge2, ENV, GEN) %>% head(3)
remove_cols(data_ge2, ENV:GEN) %>% head(3)
remove_cols(data_ge2, 1:2) %>% head(3)
remove_rows(data_ge2, 1, 2, 5, 6, 7, 8)
remove_rows(data_ge2, 1:2, 5:8)
```


## Concatenating columns
The function `concatetate()` can be used to concatenate either two columns of a data frame or a column and specified values. Return a data frame with all the original columns in .data plus the concatenated variable, after the last column.


```{r}
concatenate(data_ge, ENV, GEN)
```

If only one column is in the data, `col_1` or `col_2` must be a numeric or a character vector of length equals to the number of rows in data. Vectors of length 1 will be recycled.

```{r}
concatenate(data_ge, ENV, rep(1:14, each=30))

```

Other functions can be used to provide such vector. Here, we will replace the string "G" in the variable `GEN` with "GEN" and concatenate this column with the column `ENV`. Note that the argument `pull` into the function `replace_string()` is used to pull out the result to a vector. 

```{r}
conc_data <- 
concatenate(data_ge,
            col_1 = ENV,
            col_2 = replace_string(data_ge,
                                   GEN,
                                   replacement = "GEN_",
                                   pull = TRUE),
            new_var = ENV_GEN)
conc_data
```

To check if a column exists in a data frame, use `column_exists()`

```{r}
column_exists(conc_data, "ENV_GEN")
```

## Getting levels
To get the levels and the size of the levels of a factor, the functions `get_levels()` and `get_level_size()` can be used.

```{r}
get_levels(conc_data, ENV)
get_level_size(conc_data, ENV)
```

## Selecting numeric and non-numeric columns
Numeric columns can be selected quickly by using the function `select_numeric_cols()`. Non-numeric columns are selected with `select_non_numeric_cols()`

```{r}
select_numeric_cols(data_ge2)
select_non_numeric_cols(data_ge2)
```


# Utilities for numbers and strings
## Rounding whole data frames
The function `round_column()`round a selected column or a whole data frame to the specified number of decimal places (default 0). If no variables are informed, then all numeric variables are rounded.

```{r}
head(data_ge2)
round_column(data_ge2)
```

Alternatively, select variables to round.
```{r}
round_column(data_ge2, PH, EP, digits = 1)
```

## Extracting and replacing numbers

The functions `extract_number()`, and `replace_number()` can be used to extract or replace numbers. As an example, we will extract the number of each genotype in `data_g`. By default, the extracted numbers are put as a new variable called `new_var` after the last column of the data.

```{r}
extract_number(data_ge, GEN)
```

If the argument `drop` is set to `TRUE` then, only the new variable is kept and all others are dropped.

```{r}
extract_number(data_ge, GEN, drop = TRUE)
```

To pull out the results into a vector, use the argument `pull = TRUE`. This is particularly useful when `extract_*` or `replace_*` are used within a function like `mutate()`.


```{r}
extract_number(data_ge, GEN, pull = TRUE)
```

To replace numbers of a given column with a specified replacement, use `replace_number()`. By default, numbers are replaced with "". The argument `drop` and `pull` can also be used, as shown above.

```{r}
replace_number(data_ge, GEN)
replace_number(data_ge,
               var = REP,
               pattern = "1",
               replacement = "Rep_1",
               new_var = R_ONE)
```

## Extracting and replacing strings
The functions `extract_string()`, and `replace_string()` are used in the same context of `extract_number()`, and `replace_number()`, but for handling with strings.

```{r}
extract_string(data_ge, GEN)
```

To replace strings, we can use the function `replace_strings()`.
```{r}
replace_string(data_ge,
               var = GEN,
               new_var = GENOTYPE,
               replacement = "GENOTYPE_")
```


