---
title: "Helper functions for metan"
author: "Tiago Olivoto"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
link-citations: true
bibliography: metanref.bib 
vignette: >
  %\VignetteIndexEntry{Helper functions for metan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r global_options, include = FALSE}
knitr::opts_chunk$set(comment = "#", collapse = TRUE)
load("cvali.RData")
```


```{r warning = FALSE, message = FALSE, echo=FALSE}
library(metan)
library(DT) # Used to make the tables
# Function to make HTML tables
print_table <- function(table, rownames = FALSE, digits = 3, ...){
  df <- datatable(table, rownames = rownames, extensions = 'Buttons',
                  options = list(scrollX = TRUE, 
                                 dom = '<<t>Bp>',
                                 buttons = c('copy', 'excel', 'pdf', 'print')), ...)
  num_cols <- c(as.numeric(which(sapply(table, class) == "numeric")))
  if(length(num_cols) > 0){
    formatSignif(df, columns = num_cols, digits = digits)
  } else{
    df
  }
}

```

See the section [Rendering engine](#rendering) to know how HTML tables were generated.

# Select helpers
The package `metan` reexports the  [tidy select helpers](https://tidyselect.r-lib.org/reference/select_helpers.html) and implements own select helpers based on operations with prefixes and suffixes (`difference_var()`, `intersect_var()`, and `union_var()`), length of variable names (`width_of()`, `width_greater_than()`, and `width_less_than()`), and on case type (`lower_case_only()`, `upper_case_only()`, and `title_case_only()`).

## Variables that start with a prefix **and** ends with a suffix.

Here, we will select the variables from `data_ge2` that start with "C" and ends with "D". Just to reduce the length of outputs, only three rows are selected

```{r message=FALSE, warning=FALSE}
library(metan)
data_sel <- head(data_ge2, 3)
data_sel %>% 
  select_cols(intersect_var("C", "D")) %>% 
  print_table()
```


## Variables that start with a prefix **OR** ends with a suffix.

The following code select variables that start with "C" or ends with "D".

```{r}
data_sel %>% 
  select_cols(union_var("C", "D")) %>% 
  print_table()
```


## Variables that start with a prefix **AND NOT** ends with a suffix.

The following code select variables that start with "C" and not ends with "D".

```{r}
data_sel %>% 
  select_cols(difference_var("C", "D")) %>% 
  print_table()
```

## Selection based on length of column names.

* Select variables with an specific name length (four letters)
  
```{r}
data_sel %>% 
  select_cols(width_of(4)) %>% 
  print_table()
```

* Select variables with width less than *n*.
  
```{r}
data_sel %>% 
  select_cols(width_less_than(3)) %>% 
  print_table()
```

* Select variables with width greater than *n*.
  
```{r}
data_sel %>% 
  select_cols(width_greater_than(2)) %>% 
  print_table()
```


## Select variables by case type

Let's create data frame with 'messy' columnn names.

```{r}
df <- head(data_ge, 3)
colnames(df) <- c("Env", "gen", "Rep", "GY", "hm")
select_cols(df, lower_case_only()) %>% print_table()
select_cols(df, upper_case_only()) %>% print_table()
select_cols(df, title_case_only()) %>% print_table()
```



# Remove rows or colums wih NA values
The functions `remove_rows_na()` and `remove_rows_na()` are used to remove rows and columns with NA values, respectively.

```{r}
data_with_na <- data_g
data_with_na[c(1, 5, 10), c(3:5, 10:15)] <- NA
print_table(data_with_na)
remove_cols_na(data_with_na) %>% print_table()
remove_rows_na(data_with_na) %>% print_table()
```


# Bind cross-validation objects

```{r, eval=FALSE }


AMMI0 <- cv_ammi(data_ge, ENV, GEN, REP, GY, naxis = 0)
AMMI2 <- cv_ammi(data_ge, ENV, GEN, REP, GY, naxis = 2)
AMMI9 <- cv_ammi(data_ge, ENV, GEN, REP, GY, naxis = 9)
AMMIF <- cv_ammif(data_ge, ENV, GEN, REP, GY)
BLUP_g <- cv_blup(data_ge, ENV, GEN, REP, GY)

```


```{r, message=FALSE, fig.height = 5, fig.width = 10, fig.align = "center" }
bind1 <- bind_cv(AMMI0, AMMI2, AMMI9)
bind2 <- bind_cv(AMMI0, AMMI2, AMMI9, bind = "means")
    ```

# Split a dataframe into subsets grouped by one or more factors
## Group data and exclude all non-numeric variables
```{r, message=FALSE, fig.height = 5, fig.width = 5.5, fig.align = "center" }
g1 <- split_factors(data_ge, ENV)
is.split_factors(g1)

```


## Group data and keep all original variables
```{r, message=FALSE, fig.height = 5, fig.width = 5.5, fig.align = "center" }
g2 <- split_factors(data_ge, ENV, GEN, keep_factors = TRUE)
print_table(g2[[1]])
```


## Group a data frame using all factor variables
```{r, message=FALSE, fig.height = 5, fig.width = 5.5, fig.align = "center" }
g3 <- as.split_factors(CO2)
names(g3)
    ```


# Make a two-way table based on categorical and numerical arguments

```{r, message=FALSE }
print_table(data_ge)
matrix <- make_mat(data_ge, row = GEN, col = ENV, val = GY)
print_table(matrix, rownames = TRUE)

```

# Make upper and lower triangular matrices
```{r, message=FALSE }
cor_mat <- corr_coef(data_ge2, EP, EL, ED, EL, CD, CL)$cor

# Upper triangular
upp_mat <- make_upper_tri(cor_mat)
print_table(upp_mat, rownames = TRUE)

# Lower triangular
low_mat <- make_lower_tri(cor_mat)
print_table(low_mat, rownames = TRUE)

```

# Make a symmetric matrix

```{r, message=FALSE }
sym <- make_sym(low_mat)
print_table(sym, rownames = TRUE)

```


# Reorder a correlation matrix

Reorder the correlation matrix according to the correlation coefficient by using  hclust for hierarchical clustering order. This is useful to identify the hidden pattern in the matrix.

```{r}
print_table(cor_mat, rownames = TRUE)
reorder_cormat(cor_mat) %>% print_table(rownames = TRUE)
```


# Compute harmonic and geometric means
```{r, message=FALSE }
num <- c(1:20, 30, 50)
hmean(num)
gmean(num)

hmean(data_ge2) %>% round(2)
gmean(data_ge2, EP, EL, CL)
```



# Generate pairwise combinations of variables by applying one function to each pair
```{r, message=FALSE}
data <- data.frame(A = runif(n = 5, min = 3, max = 30),
                   B = runif(n = 5, min = 1, max = 10),
                   C = runif(n = 5, min = 9, max = 90),
                   D = runif(n = 5, min = 1, max = 90),
                   E = runif(n = 5, min = 5, max = 10))
c1 <- comb_vars(data)
print_table(c1)

c2 <- comb_vars(data, FUN = "*", order = "first")
print_table(c2)
    ```




# Combining data.frames by row, filling missing values

```{r, message=FALSE }
df1 <- data.frame(v1 = c(1, 2), v2 = c(2, 3))
df2 <- data.frame(v3 = c(4, 5))
rbind_fill(df1, df2) %>% print_table()
rbind_fill(df1, df2, fill = "NA") %>% print_table()
    ```

# Rescale a continuous vector to have specified minimum and maximum values
## Rescale a numeric vector
```{r, message=FALSE }
resca(values = c(1:5))

```

## Rescale using a data frame and select rescaled variables only
```{r, message=FALSE }
data_ge %>%
resca(GY, HM, new_min = 0, new_max = 1,  keep = FALSE) %>%
  head()%>%
  print_table()

```

## Rescale within factors
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
  data_ge2 %>% 
    select(ENV, GEN, starts_with("N"), ends_with("L")) %>%
    group_by(ENV, GEN) %>%
    summarise_all(mean) %>%
    group_by(ENV) %>%
    resca(ends_with("L")) %>%
    head(n = 13) %>%
    print_table()

```


_______________________________________________________________________________________________
# Rendering engine {#rendering}

This vignette was built with [pkgdown](https://pkgdown.r-lib.org/). All tables were produced with the package [`DT`](https://rstudio.github.io/DT/) using the following function.

```{r warning = FALSE, message = FALSE, eval=FALSE}
library(DT) # Used to make the tables
# Function to make HTML tables
print_table <- function(table, rownames = FALSE, digits = 3, ...){
  df <- datatable(table, rownames = rownames, extensions = 'Buttons',
                  options = list(scrollX = TRUE, 
                                 dom = '<<t>Bp>',
                                 buttons = c('copy', 'excel', 'pdf', 'print')), ...)
  num_cols <- c(as.numeric(which(sapply(table, class) == "numeric")))
  if(length(num_cols) > 0){
    formatSignif(df, columns = num_cols, digits = digits)
  } else{
    df
  }
}

```
