---
title: "Analyzing multienvironment trials using AMMI"
always_allow_html: yes
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: vignette
    readme: TRUE
    toc: yes
    number_sections: yes
    keep_md: TRUE
fig_caption: yes
link-citations: true
bibliography: METAABref.bib    
vignette: >
  %\VignetteIndexEntry{Analyzing multienvironment trials using AMMI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Getting started
A dataset called `data_ge` is provided to make reproducible examples. For more information, please, see `?data_ge`. Other data sets can be used provided that the following columns are in the dataset: environment, genotype, block/replicate and response variable(s).

```{r warning = FALSE, message = FALSE}
# Importing data
library(METAAB)
library(ggplot2)
library(cowplot) # used in this material to arrange the graphics
dataset = data_ge
```



# Estimating the response variable using traditional AMMI model

An interesting feature of `WAASB` package for traditional AMMI model estimation is the implementation of the S3 method `predict()`. The response variable of a two-way table (for example, the yield of *m* genotypes in *n* environments) may be estimated using the function `predict(model)`, where `model` is an object of class `WAAS.AMMI`. This estimation is based on the number of multiplicative terms declared in the function. If `naxis = 0` is declared, only the main effects (AMMI0) are considered. In this case, the estimated mean will be the estimate from OLS estimation. If `naxis = 1`, the AMMI1 (with one multiplicative term) is used for estimating the response variable. If `naxis = min(gen-1;env-1)`, the AMMIF is fitted. A summary of all possible AMMI models is presented below.

| Member of AMMI family  | Espected response of the *i*-th genotype in the *j*th environment|
|:------------------------|:------------------------------------------------------------------------------|
| AMMI0            | $\hat{y}_{ij} = \bar{y}_{i.} + \bar{y}_{.j} - \bar{y}_{..}$                   |
| AMMI1            |$\hat{y}_{ij} = \bar{y}_{i.} + \bar{y}_{.j} - \bar{y}_{..} +\lambda_1 a_{i1}t_{j1}$ |
| AMMI2            |$\hat{y}_{ij} = \bar{y}_{i.} + \bar{y}_{.j} - \bar{y}_{..} +\lambda_1 a_{i1}t_{j1}+\lambda_2 a_{i2}t_{j2}$ |
| ...              |                                                                               |
| AMMIF            |$\hat{y}_{ij} = \bar{y}_{i.} + \bar{y}_{.j} - \bar{y}_{..} +\lambda_1 a_{i1}t_{j1}+\lambda_2 a_{i2}t_{j2}+...+\lambda_p a_{ip}t_{jp}$ 


The number of axes used in the estimation must be carefully chosen. Procedures based on postdictive success, such as Gollobs's test [@Gollob:1968] or predictive success, such as cross-validation procedures [@Piepho:1994] should be used. This package provides both. `WAAS.AMMI` function compute traditional AMMI analysis showing the number of significant axes according Gollobs's test. On the other hand, `validation.AMMIF` function provides a cross-validation, estimating the RMSPD for all AMMI model family based on re-sampling procedures, considering an completely randomized desing or a randomized complete block design.

* Estimating the yield of the 10 oat cultivars in 16 environments using 5 multiplicative terms.
```{r echo = TRUE}
AMMI_model = WAAS.AMMI(dataset,
                       resp = GY,
                       gen = GEN,
                       env = ENV,
                       rep = REP,
                       verbose = FALSE)
predictoat = predict(AMMI_model, naxis = 5)

library(kableExtra)
options(digits = 4)
predictoat = predictoat[[1]][1:10,]
kable(predictoat, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                full_width = F, position = "float_left", font_size = 12)

    ```


Only the first ten values are shown. The following values are presented: **ENV** is the environment; **GEN** is the genotype; **Y** is the response variable; **resOLS** is the residual ($\hat{z}_{ij}$) estimated by the Ordinary Least Square (OLS), where $\hat{z}_{ij} = y_{ij} - \bar{y}_{i.} - \bar{y}_{.j} + \bar{y}_{ij}$; **Ypred** is the predicted value by OLS ($\hat{y}_{ij} = y_{ij} -\hat{z}_{ij}$); **ResAMMI** is the residual estimated by the AMMI model ($\hat{a}_{ij}$) considering the number of multiplicative terms informed in the function (in this case 5), where $\hat{a}_{ij} = \lambda_1\alpha_{i1}\tau_{j1}+...+\lambda_5\alpha_{i5}\tau_{j5}$; **YpredAMMI** is the predicted value by AMMI model  $\hat{ya}_{ij} = \bar{y}_{i.} + \bar{y}_{.j} - \bar{y}_{ij}+\hat{a}_{ij}$; and **AMMI0** is the predicted value when no multiplicative terms are used, i.e., $\hat{y}_{ij} = \bar{y}_{i.} + \bar{y}_{.j} - \bar{y}_{ij}$.


# Estimating the WAAS index
The `WAAS.AMMI` function computes the Weighted Average of Absolute Scores considering (i) all principal component axes that were significant ($p < 0.05$ by default); or (ii) declaring a specific number of axes to be used, according to the following equation:

$$
        WAAS_i  = 
        \sum_{k = 1}^{p} |IPCA_{ik} \times EP_k|/ \sum_{k = 1}^{p}EP_k
$$

where $WAAS_i$ is the weighted average of absolute scores of the *i*th genotype; $PCA_{ik}$ is the score of the *i*th genotype in the *k*th IPCA; and $EP_k$ is the explained variance of the *k*th IPCA for $k = 1,2,..,p$, considering *p* the number of significant PCAs, or a declared number of PCAs. The following functions may be used to do that.

## Number of axes based on F-test
In this example only IPCAs with *P*-value < 0.05 will be considered in the WAAS estimation.

```{r echo = TRUE}
# Assuming equal weights for productivity and stability (default)
WAAS1 = WAAS.AMMI(dataset,
                  resp = GY,
                  gen = GEN,
                  env = ENV,
                  rep = REP)
    ```


```{r }
# printing the WAAS object
options(digits = 3)
data = WAAS1$GY$individual$individual
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                full_width = F, position = "left", font_size = 12)

    ```


The above table shows the within-environment ANOVA considering a fixed-effect model. For each environment the Mean Squares for block, genotypes and error are shown. Estimated F-value and the probability error are also shown for block and genotype effects. Some measures of experimental precision are calculated, namelly, coefficient of variation, $CV = (\sqrt{MS_{res}}/Mean) \times 100$; the heritability, $h2 = (MS_{gen} - MS_{res})/MS_{gen}$; the accuracy of selection, $As = \sqrt{h2}$; and the coefficient of determination (R2).


```{r }
# printing the WAAS object
data = WAAS1$GY$anova
kable(data,  align  = "l", booktabs = T, format = "html", linesep = "") %>%
kable_styling(bootstrap_options = "striped", "condensed",
              position = "left", full_width = F, font_size = 12) %>%
row_spec(9, bold = T) %>%
add_indent(c(5:13))

    ```

The above table is the traditional AMMI analysis. Nine principal component axis were fitted and five were significant at 5% probability error. Based on this result, the AMMI5 model would be the best model to predict the yielding of the genotypes in the studied environments. This is confirmed by the cross-validation in the [section 3](#predictive-accuracy). The AMMI model with the smallest RMSE was the AMMI5. The predicted values (first ten observations) can be seen in [section 4](#predicting-the-yield-using-traditional-ammi-model) 

> printing the METAAB object


```{r }
options(digits = 4)
data = WAAS1$GY$model[, c(1:3,13:17, 21:22)]
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)


    ```

In this example, the scores of the nine PCA were not shown. The output generated by the `WAAS.AMMI` function shows the following results: **type**, genotype (GEN) or environment (ENV); **Code**, the code attributed to each level of the factors; **Y**, the response variable (in this case the grain yield); **WAAS** the weighted average of the absolute scores, estimated with all PCA axes with *P*-value $\le$ 0.05; **PctWAAS** and **PctResp** that are the percentage values for the WAAS and Y, respectively; **OrResp** and **OrWAAS** that are the ranks attributed to the genotype and environment regarding the Y or WAAS, respectively; **WAASY** is the weighted average of absolute scores and response variable. In this case, considering equal weights for PctResp and PctWAAS, the WAASY for G1 is estimated by: $WAAS_{G1} = [(86.32\times50)+(98.88\times50)]/50+50 = 92.60$. Then the **OrWAASY* is the rank for the WAASY value. The genotype (or environment) with the largest WAASY value has the first ranked. See [Estimating the WAASBY index](#estimating-the-waasby-index) for a detailed explanation.

In this example, the scores of the nine PCAs were not shown. The output generated by the `WAAS.AMMI` function shows the following results: **type**, genotype (GEN) or environment (ENV); **Code**, the code attributed to each level of the factors; **Y**, the response variable (in this case the grain yield); **WAAS** the weighted average of the absolute scores, estimated with all significant IPCA; **PctWAAS** and **PctResp**, the rescaled variable for the WAAS and Y, respectively; **OrResp** and **OrWAAS**, the ranks attributed to the genotype and environment regarding the Y or WAAS, respectively; **WAASY** is the simultanteous selection index that weights between response variable and stability. In this case, considering equal weights for PctResp and PctWAAS, the WAASY for G1 would be then: $WAASY_{G1} = [(22.043\times50)+(100\times50)]/50+50 = 92.60$, which, with equal weights is equivalente to the aritmetic mean of pctResp and PctWAAS. Then the *OrWAASY* is the rank for the WAASY value. The genotype (or environment) with the largest WAASY value has the first ranked. Please, refer to [Estimating the WAASBY index](#estimating-the-waasby-index) for a detailed explanation.


## Number of axes declared manually
The second option to compute the WAAS is by manually declaring a specific number of multiplicative terms. In this case, the number of terms declared is used independently of its significance. Let us, for the moment, assume that after a cross-validation procedure the AMMI7 was the most predictively accurate AMMI model and the researcher will use this model. The additional argument `naxis` in the function `WAAS.AMMI` is then used to overwrite the default chose of significant terms. 

```{r echo = TRUE}
WAAS2 = WAAS.AMMI(dataset,
                  resp = GY,
                  gen = GEN,
                  env = ENV,
                  rep = REP,
                  naxis = 7,
                  verbose = FALSE)

    ```

The only difference in this output compared to those from  [section 5.1](#assuming-a-given-probability-error-for-chosing-the-number-of-axes) is that here we declared that seven PCA axes should be used for computing the WAAS value. Thus, only the values of WAAS, OrWAAS, WAASY and OrWAASY may have significant changes.


# Other AMMI-based stability indexes
The following AMMI-based stability indexes tested in the article may be computed using the function `AMMI_indexes()`:
 
 * **AMMI stability value, ASV, [@Purchase2000].**

$$
ASV = \sqrt {{{\left[ {\frac{{IPCA{1_{ss}}}}{{IPCA{2_{ss}}}} \times \left( {IPCA{1_{score}}} \right)} \right]}^2} + {{\left( {IPCA{2_{score}}} \right)}^2}}
$$

* **Sums of the absolute value of the IPCA scores**

$$
SIP{C_i} = \sum\nolimits_{k = 1}^P {\left| {\mathop {\lambda }\nolimits_k^{0.5} {a_{ik}}} \right|}
$$

* **Averages of the squared eigenvector values**

$$
E{V_i} = \sum\nolimits_{k = 1}^P {\mathop a\nolimits_{ik}^2 } /P
$$
described by @Sneller1997, where *P* is the number of IPCA retained via F-tests;

* **absolute value of the relative contribution of IPCAs to the interaction [@Zali2012].**

$$
Z{a_i} = \sum\nolimits_{k = 1}^P {{\theta _k}{a_{ik}}}
$$

where ${\theta _k}$ is the percentage sum of squares explained by the *k*-th IPCA. Simultaneous selection indexes (ssi), are computed by summation of the ranks of the ASV, SIPC, EV and Za indexes and the ranks of the mean yields [@Farshadfar2008], which results in ssiASV, ssiSIPC, ssiEV, and ssiZa, respectively.


```{r }
stab_indexes = AMMI_indexes(AMMI_model)
kable(stab_indexes, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

    ```


## Biplots

Provided that an object of class "WAAS.AMMI" is available in the global environment, the graphics may be obtained using the function `plot.scores()`. To do that, we will revisit the previusly fitted model `AMMI_model` . Please, refer to `?plot.scores` for more details. Four types of graphics can be generated: 1 = $PC1 \times PC2$;  2 = $GY \times PC1$; 3 = $GY \times WAASB$; and 4 = a graphic with nominal yield as a function of the environment PCA1 scores.

### biplot type 1: PC1 x PC2

```{r echo = TRUE, fig.height = 5, fig.width = 10, fig.align = "center", message = F, warning = F}
library(cowplot)
p1 = plot.scores(AMMI_model$GY, type = 1)
p2 = plot.scores(AMMI_model$GY,
                 type = 1,
                 polygon = TRUE,
                 col.gen = "black",
                 col.env = "gray70",
                 col.segm.env = "gray70",
                 axis.expand = 1.5)
plot_grid(p1, p2, labels = c("p1","p2"))
```


### biplot type 2: GY x PC1

```{r echo = TRUE, fig.height = 5, fig.width = 10, fig.align = "center", message = F, warning = F}

p3 = plot.scores(AMMI_model$GY, type = 2)
p4 = plot.scores(AMMI_model$GY, type = 2,
                 col.segm.env = "transparent") +
                 theme_gray() +
                 theme(legend.position = c(0.1, 0.9),
                       legend.background = element_rect(fill = NA))

plot_grid(p3, p4, labels = c("p3","p4"))
```

### biplot type 3: GY x WAAS

The quadrants proposed in the following biplot represent the four classifications proposed here regarding the joint interpretation of productivity and stability. The genotypes or environments included in quadrant I can be considered unstable genotypes or environments with high discrimination ability, and with productivity below the grand mean. In quadrant II are included unstable genotypes, although with productivity above the grand mean. The environments included in this quadrant deserve special attention since, in addition to providing high magnitudes of the response variable, they present a good discrimination ability. Genotypes within quadrant III have low productivity, but can be considered stable due to the lower values of WAASB. The lower this value, the more stable the genotype can be considered. The environments included in this quadrant can be considered as poorly productive and with low discrimination ability. The genotypes within the quadrant IV are higly productive and broadly adapted due to the high magnitude of the response variable and high stability performance (lower values of WAASB).    

```{r echo = TRUE, fig.height = 5, fig.width = 10, fig.align = "center", message=F, warning=F}

p5 = plot.scores(AMMI_model$GY, type = 3)
p6 = plot.scores(AMMI_model$GY, type = 3,
                 x.lab = "My customized x label",
                 size.shape = 3,
                 size.tex.pa = 2,
                 x.lim = c(1.2, 4.7),
                 x.breaks = seq(1.5, 4.5, by = 0.5)) + 
                 theme(legend.position = c(0.1, 0.9))
plot_grid(p5, p6, labels = c("p5","p6"))
```



### biplot type 4 : nominal yield and environment IPCA1

```{r echo = TRUE, fig.height = 5, fig.width = 5.5, fig.align = "center", message=F, warning=F}

plot.scores(AMMI_model$GY,
            type = 4, size.tex.pa = 1.5)


```


# Estimating the WAASY index
The function `WAASratio.AMMI` considers both stability (WAAS) and productivity for genotype ranking considering the following model:

$$
WAASY{_i} = \frac{{\left( {r{G_i} \times {\theta _Y}} \right) + \left( {r{W_i} \times {\theta _S}} \right)}}{{{\theta _Y} + {\theta _S}}}
$$

where $WAASY_i$ is the superiority index for the *i*-th genotype that weights between performance and stability; $rG_i$ and $rW_i$ are the rescaled values (0-100) for GY and WAASB, respectively;  $\theta _Y$ and $\theta_S$ are the weights for GY and WAASB, respectively.

In the following example we will assume that we want to obtain the ranks changing the WAAS/GY ratio in 10% each scenario and to plot the WAASBY values considering a WAAS/GY ratio equal 30/70. **Important! THE ARGUMENT `saveWAASY` MUST BE DIVISIBLE BY `increment`!**

```{r echo = TRUE}
WAASratio = WAASBYratio(dataset,
                          resp = GY,
                          gen = GEN,
                          env = ENV,
                          rep = REP,
                          increment = 50,
                          saveWAASY = 50)

```


## Printing the model outputs

* WAASY values

```{r }
options(digits = 4)
data = WAASratio$WAASY
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

```

In this output, **PesResp** and **PesWAAS** are the weights attributed to response variable and stability, respecively, to compute de WAASY values.

* Genotype ranking for each scenario of WAASBY/GY ratio.

```{r }

options(digits = 4)
data = WAASratio$hetcomb
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

```

* Genotype ranking depending on the number of multiplicative terms used to estimate the WAASB index.

```{r }

options(digits = 4)
data = WAASratio$hetdata
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

```


## Plotting the WAASY values

```{r echo = TRUE, fig.height = 4, fig.width = 10, fig.align = "center", message = F, warning = F}
library(ggplot2)
p1 = plot.WAASBY(WAASratio)
p2 = plot.WAASBY(WAASratio, col.shape = c("gray20", "gray80"))
plot_grid(p1, p2, labels = c("p1", "p2"))
    ```


## Plotting the heat map graphics
The first type of heatmap shows the genotype ranking depending on the number of principal component axes used for estimating the WAASB index. An euclidean distance-based dendrogram is used for grouping the genotype ranking for both genotypes and principal component axes. The second type of heatmap shows the genotype ranking depending on the WAASB/GY ratio. The ranks obtained with a ratio of 100/0 considers exclusively the stability for genotype ranking. On the other hand, a ratio of 0/100 considers exclusively the productivity for genotype ranking. Four clusters are estimated (1) unproductive and unstable genotypes; (2) productive, but unstable genotypes; (3) stable, but unproductive genotypes; and (4), productive and stable genotypes.



### Ranks of genotypes depending on the number of PCA used to estimate the WAAS
```{r echo = TRUE, fig.height = 5, fig.width = 5.5, fig.align = "center", message = F, warning = F}

plot(WAASratio, type = 1)


    ```

### Ranks of genotypes depending on the WAAS/GY ratio
```{r echo = TRUE, fig.height = 5, fig.width = 5.5, fig.align = "center", message = F, warning = F}

plot(WAASratio, type = 2)


```


# References

