---
title: "Utilities for data manipulation"
author: "Tiago Olivoto"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
link-citations: true
bibliography: metanref.bib 
vignette: >
  %\VignetteIndexEntry{Utilities for data manipulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r global_options, include = FALSE}
knitr::opts_chunk$set(comment = "#", collapse = TRUE)
```


```{r warning = FALSE, message = FALSE, echo=FALSE}
library(metan)
library(DT) # Used to make the tables
# Function to make HTML tables
print_table <- function(table, rownames = FALSE, digits = 3, ...){
datatable(table, rownames = rownames, extensions = 'Buttons',
          options = list(scrollX = TRUE, 
                         dom = '<<t>Bp>',
                         buttons = c('copy', 'excel', 'pdf', 'print')), ...) %>%
    formatSignif(columns = c(as.numeric(which(sapply(table, class) == "numeric"))), digits = digits)}

```


# Utilities for rows and columns
## Add columns and rows
The functions `add_cols()` and `add_rows()` can be used to add columns and rows, respectively to a data frame.

```{r message=FALSE, warning=FALSE}
library(metan)
add_cols(data_ge,
         ROW_ID = 1:420) %>% 
  print_table()

```

It is also possible to add a column based on existing data. Note that the arguments `.after` and `.before` are used to select the position of the new column(s). This is particularly useful to put variables of the same category together.

```{r}
add_cols(data_ge,
         GY2 = GY^2,
         .after = "GY") %>% 
  print_table()
```

## Select or remove columns and rows
The functions `select_cols()` and `select_rows()` can be used to select columns and rows, respectively from a data frame.

```{r}
select_cols(data_ge2, ENV, GEN) %>% print_table()
select_rows(data_ge2, 1:3)  %>% print_table()
```

Numeric columns can be selected quickly by using the function `select_numeric_cols()`. Non-numeric columns are selected with `select_non_numeric_cols()`

```{r}
select_numeric_cols(data_ge2) %>% print_table()
select_non_numeric_cols(data_ge2) %>% print_table()
```

We can select the first or last columns quickly with `select_first_col()` and `select_last_col()`, respectively.

```{r}
select_first_col(data_ge2) %>% print_table()
select_last_col(data_ge2) %>% print_table()
```


To remove columns or rows, use `remove_cols()` and `remove_rows()`.
```{r}
remove_cols(data_ge2, ENV, GEN)  %>% print_table()
remove_rows(data_ge2, 1:2, 5:8) %>% print_table()
```


## Concatenating columns
The function `concatetate()` can be used to concatenate multiple columns of a data frame. It return a data frame with all the original columns in `.data` plus the concatenated variable, after the last column (by default), or at any position when using the arguments `.before` or `.after`.


```{r}
concatenate(data_ge, ENV, GEN, REP, .after = "REP") %>% print_table()
```

To drop the existing variables and keep only the concatenated column, use the argument `drop = TRUE`. To use `concatenate()` within a given function like `add_cols()` use the argument `pull = TRUE` to pull out the results to a vector.
```{r}
concatenate(data_ge, ENV, GEN, REP, drop = TRUE) %>% head()
concatenate(data_ge, ENV, GEN, REP, pull = TRUE) %>% head()
```


To check if a column exists in a data frame, use `column_exists()`

```{r}
column_exists(data_ge, "ENV")
```

## Getting levels
To get the levels and the size of the levels of a factor, the functions `get_levels()` and `get_level_size()` can be used.

```{r}
get_levels(data_ge, ENV)
get_level_size(data_ge, ENV)
```


# Utilities for numbers and strings
## Rounding whole data frames
The function `round_cols()`round a selected column or a whole data frame to the specified number of decimal places (default 0). If no variables are informed, then all numeric variables are rounded.

```{r}
round_cols(data_ge2) %>% print_table()
```

Alternatively, select variables to round.
```{r}
round_cols(data_ge2, PH, EP, digits = 1)  %>% print_table()
```

## Extracting and replacing numbers

The functions `extract_number()`, and `replace_number()` can be used to extract or replace numbers. As an example, we will extract the number of each genotype in `data_g`. By default, the extracted numbers are put as a new variable called `new_var` after the last column of the data.

```{r}
extract_number(data_ge, GEN, .after = "GEN") %>% print_table()
```

If the argument `drop` is set to `TRUE` then, only the new variable is kept and all others are dropped.

```{r}
extract_number(data_ge, GEN, drop = TRUE) %>% print_table()
```

To pull out the results into a vector, use the argument `pull = TRUE`. This is particularly useful when `extract_*` or `replace_*` are used within a function like `add_cols()`.


```{r}
extract_number(data_ge, GEN, pull = TRUE)
```

To replace numbers of a given column with a specified replacement, use `replace_number()`. By default, numbers are replaced with "". The argument `drop` and `pull` can also be used, as shown above.

```{r}
replace_number(data_ge, GEN) %>% print_table()
replace_number(data_ge,
               var = REP,
               pattern = "1",
               replacement = "Rep_1",
               new_var = R_ONE,
               .after = "REP") %>% 
  print_table()
```

## Extracting, replacing, and removing strings
The functions `extract_string()`, and `replace_string()` are used in the same context of `extract_number()`, and `replace_number()`, but for handling with strings.

```{r}
extract_string(data_ge, GEN) %>% print_table()
```

To replace strings, we can use the function `replace_strings()`.
```{r}
replace_string(data_ge,
               var = GEN,
               new_var = GENOTYPE,
               replacement = "GENOTYPE_",
               .after = "GEN") %>% 
  print_table()
```

To remove all strings of a data frame, use `remove_strings()`.
```{r}
remove_strings(data_ge) %>% print_table()
```

## Tidy strings
The function `tidy_strings()` tidy up characters strings, non-numeric columns, or any selected columns in a data frame by putting all word in upper case, replacing any space, tabulation, punctuation characters by `'_'`, and putting `'_'` between lower and upper cases. Consider the following character strings: `messy_env` by definition should represent a unique level of the factor environment (environment 1). `messy_gen` shows six genotypes, and `messy_int` represents the interaction of such genotypes with environment 1.

```{r}
messy_env <- c("ENV 1", "Env   1", "Env1", "env1", "Env.1", "Env_1")
messy_gen <- c("GEN1", "gen 2", "Gen.3", "gen-4", "Gen_5", "GEN_6")
messy_int <- c("Env1Gen1", "Env1_Gen2", "env1 gen3", "Env1 Gen4", "ENV_1GEN5", "ENV1GEN6")

```

These character vectors are visually messy. Let's tidy them.
```{r}
tidy_strings(messy_env)
tidy_strings(messy_gen)
tidy_strings(messy_int)
```


`tidy_strings()` works also to tidy a whole data frame or specific columns. Let's create a 'messy' data frame in the context of plant breeding trials.

```{r}
library(tibble)
df <- tibble(Env = messy_env,
             gen = messy_gen,
             Env_GEN = interaction(Env, gen),
             y = rnorm(6, 300, 10))
df %>% print_table()
tidy_strings(df) %>% print_table()
tidy_strings(df, gen) %>% print_table()
```


_______________________________________________________________________________________________
**Rendering engine**

This vignette was built with [pkgdown](https://pkgdown.r-lib.org/). All tables were produced with the package [`DT`](https://rstudio.github.io/DT/) using the following function.

```{r warning = FALSE, message = FALSE, eval=FALSE}
library(metan)
library(DT) # Used to make the tables
# Function to make HTML tables
print_table <- function(table, rownames = FALSE, digits = 3, ...){
datatable(table, rownames = rownames, extensions = 'Buttons',
          options = list(scrollX = TRUE, 
                         dom = '<<t>Bp>',
                         buttons = c('copy', 'excel', 'pdf', 'print')), ...) %>%
    formatSignif(columns = c(as.numeric(which(sapply(table, class) == "numeric"))), digits = digits)}

```

