---
title: "Analyzing multienvironment trials using BLUP"
always_allow_html: yes
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: vignette
    readme: TRUE
    toc: yes
    number_sections: yes
    keep_md: TRUE
fig_caption: yes
link-citations: true
bibliography: METAABref.bib  
vignette: >
  %\VignetteIndexEntry{Analyzing multienvironment trials using BLUP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Getting started
A dataset called `data_ge` is provided to make reproducible examples. For more information, please, see `?data_ge`. Other data sets can be used provided that the following columns are in the dataset: environment, genotype, block/replicate and response variable(s).

```{r warning = FALSE, message = FALSE}
# Importing data
library(METAAB)
library(ggplot2)
library(cowplot)
library(kableExtra)
dataset = data_ge
```


# Estimating the WAASB (based on SVD of BLUP-interaction effects)

The `WAASB()` function computes the weighted average of absolute scores considering all possible IPCA from the Singular Value Decomposition of the BLUPs for genotype-vs-environment interaction effects obtained by an Linear Mixed-effect Model, as follows:

$$
        WAASB_i  = 
        \sum_{k = 1}^{p} |IPCA_{ik} \times EP_k|/ \sum_{k = 1}^{p}EP_k
$$

where $WAASB_i$ is the weighted average of absolute scores of the *i*th genotype; $IPCA_{ik}$ is the scores of the *i*th genotype in the *k*th IPCA; and $EP_k$ is the explained variance of the *k*th PCA for $k = 1,2,..,p$, $p = min(G-1;E-1)$.


```{r echo = TRUE}
# Assuming equal weights for productivity and stability (default)
WAASB = WAASB(dataset,
              resp = GY,
              gen = GEN,
              env = ENV,
              rep = REP)

    ```


## Diagnostic plot for residuals
The function `autoplot()` is used to generate diagnostic plots of residuals of the model. The normality of the random effects of genotype and interaction effects may be also obtained by using `type = "re"`.
```{r echo = TRUE, fig.width=7, fig.height=7, fig.align="center"}
library(ggplot2)
autoplot(WAASB$GY)
    ```


## Printing the model outputs

### Likelihood Ratio Tests

```{r warning=F, message=F}
options(digits = 5)
data = WAASB$GY$LRT
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)
    ```


The output `LRT` contains the Likelihood Ratio Tests for genotype and genotype-vs-environment random effects.


### Variance components and genetic parameters

```{r }
options(digits = 7)
data = WAASB$GY$ESTIMATES
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

    ```


In the output `ESTIMATES`, beyond the variance components for the declared random effects, some important parameters are also shown. **Heribatility** is the broad-sense heritability, $\mathop h\nolimits_g^2$, estimated by $\mathop h\nolimits_g^2  = \mathop {\hat\sigma} \nolimits_g^2 /\left( {\mathop {\hat\sigma} \nolimits_g^2  + \mathop {\hat\sigma} \nolimits_i^2  + \mathop {\hat\sigma} \nolimits_e^2 } \right)$ where $\mathop {\hat\sigma} \nolimits_g^2$ is the genotypic variance; $\mathop {\hat\sigma} \nolimits_i^2$ is the genotype-by-environment interaction variance; and $\mathop {\hat\sigma} \nolimits_e^2$ is the residual variance. **GEIr2** is the coefficient of determination of the interaction effects, $\mathop r\nolimits_i^2$, estimated by $\mathop r\nolimits_i^2  = \mathop {\hat\sigma} \nolimits_i^2 /\left( {\mathop {\hat\sigma} \nolimits_g^2  + \mathop {\hat\sigma} \nolimits_i^2  + \mathop {\hat\sigma} \nolimits_e^2 } \right)$; **Heribatility of means** is the heribability on the mean basis, $\mathop h\nolimits_{gm}^2$, estimated by $\mathop h\nolimits_{gm}^2  = \mathop {\hat\sigma} \nolimits_g^2 /\left[ {\mathop {\hat\sigma} \nolimits_g^2  + \mathop {\hat\sigma} \nolimits_i^2 /e + \mathop {\hat\sigma} \nolimits_e^2 /\left( {eb} \right)} \right]$, where *e* and *b* are the number of environments and blocks, respectively; **Accuracy** is the accuracy of selection, *Ac*, estimated by $Ac = \sqrt{\mathop h\nolimits_{gm}^2}$ ;   **rge** is the genotype-environment correlation,  $\mathop r\nolimits_{ge}$, estimated by $\mathop r\nolimits_{ge} = \mathop {\hat\sigma} \nolimits_g^2 /\left({\mathop {\hat\sigma} \nolimits_g^2  + \mathop {\hat\sigma} \nolimits_i^2} \right)$; **CVg** is the the genotypic coefficient of variation, estimated by \(\left( {\sqrt {\mathop {\hat \sigma }\nolimits_g^2 } /\mu } \right) \times 100\) , where $\mu$ is the grand mean;  **CVr** is the residual coefficient of variation, estimated by $\left( {\sqrt {\mathop {\hat \sigma }\nolimits_e^2 } /\mu } \right) \times 100$ ; **CV ratio** is the ratio between genotypic and residual coefficient of variation.


### Some useful information

```{r }
options(digits = 4)
data = WAASB$GY$Details
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

    ```


The following pieces of information are provided in `Details` output. **WgtResponse** is the weight for the response variable in estimating WAASB; **WgtWAAS** is the weight for stability; **Ngen** is the number of genotypes; **Nenv** is the number of environments; **OVmean** is the overall mean; **Min** is the minimum value observed (returning the genotype and environment); **Max** is the maximum observed; **MinENV** is the environment with the lower mean; **MaxENV** is the environment with the largest mean observed; **MinGEN** is the genotype with the lower mean; **MaxGEN** is the genotype with the largest mean.    

### The WAASB object

```{r }
options(digits = 4)
data = WAASB$GY$model[, c(1:3,13:17, 21:22)]
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

    ```

This output generated by the `WAASB` function is very similar to those shown in the [sections 5.1](#assuming-a-given-probability-error-for-chosing-the-number-of-axes) and [5.2](#declaring-a-specific-number-of-axis-to-be-used). The main difference here, is that the singular value decomposition is based on the BLUP interaction effect matrix and the **WAASB** in this output is the weighted average of the absolute scores, estimated with all estimated PCA axes, where instead **WAAS** that is estimated considering only PCA axes with *P*-value $\le$ 0.05.

### BLUP for genotypes

```{r }
options(digits = 4)
data = WAASB$GY$blupGEN[1:10,]
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

    ```


### Plotting the BLUP for genotypes

```{r echo = TRUE, fig.height = 5, fig.width = 10, fig.align = "center", message = F, warning = F}
# No file exported
p1 = plot.blup(WAASB$GY)
p2 = plot.blup(WAASB$GY, 
               col.shape  =  c("gray20", "gray80")) + coord_flip()
plot_grid(p1, p2,
          labels = c("p1", "p2"))

```

This output shows the predicted means for genotypes. **BLUPg** is the genotypic effect $(\hat{g}_{i})$ estimated by $\hat{g}_{i} = h_g^2(\bar{y}_{i.}-\bar{y}_{..})$ where $h_g^2$ is the shrinkage effect for genotype. **Predicted** is the predicted mean estimated by $\hat{g}_{i}+\mu$ where $\mu$ is the grand mean. **LL** and **UL** are the lower and upper limits, respectively, estimated by $(\hat{g}_{i}+\mu)\pm{CI}$. $CI$ is the confidence interval for BLUP prediction assuming a given probability error, where  $CI = t\times\sqrt{((1-Ac)\times{\mathop \sigma \nolimits_g^2)}}$ where $t$ is the Student's *t* value for a two-tailed t test at a given probability error; $Ac$ is the accuracy of selection and $\mathop \sigma \nolimits_g^2$ is the genotypic variance.

### BLUP for genotypes X environment combination

```{r }
options(digits = 4)
data = WAASB$GY$BLUPgge[1:10,]
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

    ```

This output shows the predicted means for each genotype and environment combination. **BLUPg** is the genotypic effect described above. **BLUPge** is the genotypic effect of the *i*th genotype in the *j*th environment $(\hat{g}_{ij})$ estimated by $\hat{g}_{ij} = h_g^2(\bar{y}_{i.}-\bar{y}_{..})+h_{ge}^2(y_{ij}-\bar{y}_{i.}-\bar{y}_{.j}+\bar{y}_{..})$, where $h_{ge}^2$ is the shrinkage effect for the genotype-by-environment interaction; **BLUPg+ge** is $BLUP_g+BLUP_{ge}$; **Predicted** is the predicted mean ($\hat{y}_{ij}$) estimated by $\hat{y}_{ij} = \bar{y}_{.j}+BLUP_{g+ge}$.

## Eigenvalues from the SVD

```{r }
options(digits = 4)
data = WAASB$GY$PCA
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

    ```


```{r echo = TRUE, fig.height = 5, fig.width = 5.5, fig.align = "center", message = F, warning = F}

plot.eigen(WAASB$GY, size.lab = 14, size.tex.lab = 14)

```

The above output shows the eigenvalues and the proportion of variance explained by each principal component axis of the BLUP interaction effects matrix.

### Phenotypic means

```{r }
options(digits = 4)
data = WAASB$GY$MeansGxE[1:10,]
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)
    ```

In this output, *Y* is the phenotypic mean for each genotype and environment combination ($y_{ij}$), estimated by $y_{ij} = \sum_k{y_{ij}}/B$ with $k = 1,2,...B$.


## Biplots

We will show how biplots may be obtained for both traditional AMMI model, fitted by the function `WAAS.AMMI()` and the mixed-effect model fitted by the function `WAASB()`. Provided that an object of class "WAAS.AMMI" or "WAASB" is available in the global environment, the graphics may be obtained using the function `plot.scores()`. To do that, we will revisit the previusly fitted model `WAASB` . Please, refer to `?plot.scores` for more details. Four types of graphics can be generated: 1 = $PC1 \times PC2$;  2 = $GY \times PC1$; 3 = $GY \times WAASB$; and 4 = a graphic with nominal yield as a function of the environment PCA1 scores.

### biplot type 1: PC1 x PC2

```{r echo = TRUE, fig.height = 5, fig.width = 10, fig.align = "center", message = F, warning = F}
library(cowplot)
p1 = plot.scores(WAASB$GY, type = 1)
p2 = plot.scores(WAASB$GY,
                 type = 1,
                 polygon = TRUE,
                 col.gen = "black",
                 col.env = "gray70",
                 col.segm.env = "gray70",
                 axis.expand = 1.5)
plot_grid(p1, p2, labels = c("p1","p2"))
```


### biplot type 2: GY x PC1

```{r echo = TRUE, fig.height = 5, fig.width = 10, fig.align = "center", message = F, warning = F}

p3 = plot.scores(WAASB$GY, type = 2)
p4 = plot.scores(WAASB$GY, type = 2,
                 col.segm.env = "transparent") +
                 theme_gray() +
                 theme(legend.position = c(0.1, 0.9),
                       legend.background = element_rect(fill = NA))

plot_grid(p3, p4, labels = c("p3","p4"))
```

### biplot type 3: GY x WAASB

The quadrants proposed in the following biplot represent the four classifications proposed here regarding the joint interpretation of productivity and stability. The genotypes or environments included in quadrant I can be considered unstable genotypes or environments with high discrimination ability, and with productivity below the grand mean. In quadrant II are included unstable genotypes, although with productivity above the grand mean. The environments included in this quadrant deserve special attention since, in addition to providing high magnitudes of the response variable, they present a good discrimination ability. Genotypes within quadrant III have low productivity, but can be considered stable due to the lower values of WAASB. The lower this value, the more stable the genotype can be considered. The environments included in this quadrant can be considered as poorly productive and with low discrimination ability. The genotypes within the quadrant IV are higly productive and broadly adapted due to the high magnitude of the response variable and high stability performance (lower values of WAASB).    

```{r echo = TRUE, fig.height = 5, fig.width = 10, fig.align = "center", message=F, warning=F}

p5 = plot.scores(WAASB$GY, type = 3)
p6 = plot.scores(WAASB$GY, type = 3,
                 x.lab = "My customized x label",
                 size.shape = 3,
                 size.tex.pa = 2,
                 x.lim = c(1.2, 4.7),
                 x.breaks = seq(1.5, 4.5, by = 0.5)) + 
                 theme(legend.position = c(0.1, 0.9))
plot_grid(p5, p6, labels = c("p5","p6"))
```



### biplot type 4 : nominal yield and environment IPCA1

```{r echo = TRUE, fig.height = 5, fig.width = 5.5, fig.align = "center", message=F, warning=F}

plot.scores(WAASB$GY,
            type = 4, size.tex.pa = 1.5)


```


# Estimating the WAASBY index
The function `WAASBratio` considers both stability (WAASB) and productivity for genotype ranking considering the following model:

$$
WAASB{Y_i} = \frac{{\left( {r{G_i} \times {\theta _Y}} \right) + \left( {r{W_i} \times {\theta _S}} \right)}}{{{\theta _Y} + {\theta _S}}}
$$

where $WAASBY_i$ is the superiority index for the *i*-th genotype that weights between performance and stability; $rG_i$ and $rW_i$ are the rescaled values (0-100) for GY and WAASB, respectively;  $\theta _Y$ and $\theta_S$ are the weights for GY and WAASB, respectively.

This function provides the option of attributing weights for stability and productivity in genotype ranking. This is important depending on the goal of a selection strategy or cultivar reccomendation. For example, if the goal is to select a genotype with high yield (independently of the stability performance), that genotype with the first rank in a WAASB/GY = 0/100 ratio should be selected. The reciprocal is true. Aiming at selecting high-stable genotypes (independently of the productivity), that genotype with the first rank in a WAASB/GY = 100/0 ratio should be selected. By default, the increment on the WAASB/GY ratio is equal to 5 and the WAASBY values are saved when the WAASB/GY ratio is equal to 50/50. Thus, twenty one different scenarios are computed, and for each scenario, the genotypes are ranked based on the WAASBY values.

In the following example we will assume that we want to obtain the ranks changing the WAASB/GY ratio in 10% each scenario and to plot the WAASBY values considering a WAASB/GY ratio equal 30/70. **Important! THE ARGUMENT `saveWAASY` MUST BE DIVISIBLE BY `increment`!**

```{r echo = TRUE}
WAASBYratio = WAASBYratio(dataset,
                          resp = GY,
                          gen = GEN,
                          env = ENV,
                          rep = REP,
                          increment = 50,
                          saveWAASY = 50)

```

This procedure can also be used with the traditional AMMI analysis. This approach is easily implemented using the `WAASratio.AMMI` function shown in the following example. 

```{r echo = TRUE, eval=FALSE}
WAASBYratio2 = WAASratio.AMMI(dataset,
                              resp = GY,
                              gen = GEN,
                              env = ENV,
                              rep = REP,
                              increment = 50,
                              saveWAASY = 50)

```


## Printing the model outputs

* WAASBY values

```{r }
options(digits = 4)
data = WAASBYratio$WAASY
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

```

In this output, **PesResp** and **PesWAAS** are the weights attributed to response variable and stability, respecively, to compute de WAASY values.

* Genotype ranking for each scenario of WAASBY/GY ratio.

```{r }

options(digits = 4)
data = WAASBYratio$hetcomb
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

```

* Genotype ranking depending on the number of multiplicative terms used to estimate the WAASB index.

```{r }

options(digits = 4)
data = WAASBYratio$hetdata
kable(data, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

```


## Plotting the WAASBY values

```{r echo = TRUE, fig.height = 4, fig.width = 10, fig.align = "center", message = F, warning = F}
library(ggplot2)
p1 = plot.WAASBY(WAASBYratio)
p2 = plot.WAASBY(WAASBYratio, col.shape = c("gray20", "gray80"))
plot_grid(p1, p2, labels = c("p1", "p2"))
    ```


## Plotting the heat map graphics
The first type of heatmap shows the genotype ranking depending on the number of principal component axes used for estimating the WAASB index. An euclidean distance-based dendrogram is used for grouping the genotype ranking for both genotypes and principal component axes. The second type of heatmap shows the genotype ranking depending on the WAASB/GY ratio. The ranks obtained with a ratio of 100/0 considers exclusively the stability for genotype ranking. On the other hand, a ratio of 0/100 considers exclusively the productivity for genotype ranking. Four clusters are estimated (1) unproductive and unstable genotypes; (2) productive, but unstable genotypes; (3) stable, but unproductive genotypes; and (4), productive and stable genotypes.



### Ranks of genotypes depending on the number of PCA used to estimate the WAASB
```{r echo = TRUE, fig.height = 5, fig.width = 5.5, fig.align = "center", message = F, warning = F}

plot(WAASBYratio, type = 1)


    ```

### Ranks of genotypes depending on the WAASB/GY ratio
```{r echo = TRUE, fig.height = 5, fig.width = 5.5, fig.align = "center", message = F, warning = F}

plot(WAASBYratio, type = 2)


```


# Others BLUP-based stability indexes

@ColombariFilho2013 have shown the use of three BLUP-based indexes for selecting genotypes with performance and stability. The first is the harmonic mean of genotypic values -or BLUPS- (HMGV) a stability index that considers the genotype with the highest harmonic mean across environments as the most stable, as follows:

$$
HMGV_i = \frac{1}{e}\sum\limits_{j = 1}^e {\frac{1}{{BLUP_{ij}}}}
$$

The second is the relative performance of genotypic values (RPGV), an adaptability index estimated as follows:

$$
RPGV_i = \frac{1}{e}{\sum\limits_{j = 1}^e {BLUP_{ij}} /\mathop \mu \nolimits_j }
$$

The third and last is the harmonic mean of relative performance of genotypic values (HMRPGV), a simultaneous selection index for stability, adaptability and mean performance, estimated as follows:

$$
HMRPGV_i = \frac{1}{e}\sum\limits_{j = 1}^e {\frac{1}{{BLUP_{ij}/{\mu _j}}}}
$$


```{r }
res_inde = Resende_indexes(WAASB)
kable(res_inde, "html") %>%
  kable_styling(bootstrap_options = "striped", "condensed",
                position = "left", full_width = F, font_size = 12)

    ```




# References

