---
title: "Parametric and non-parametric stability statistics"
always_allow_html: yes
output: rmarkdown::html_vignette
fig_caption: yes
link-citations: true
bibliography: METAABref.bib    
vignette: >
  %\VignetteIndexEntry{Parametric and non-parametric stability statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Getting started

In this section, we will use the data in `data_ge2`. For more information, please, see `?data_ge2`. Other data sets can be used provided that the following columns are in the dataset: environment, genotype, block/replicate and response variable(s).

```{r, fig.width=5, fig.height=4, warning = FALSE, message = FALSE}
library(METAAB)
library(cowplot) # used to arrange the graphics
library(kableExtra) # Used to make the tables
dataset = data_ge2
str(dataset)
```

# Graphic visualization of the interaction
The function `ge_plot()` may be used to visualize the genotype's performance across the environments. The black diamond shows the mean of each environment.
```{r, fig.width=10, fig.height=5, warning = FALSE, message = FALSE}
p1 = ge_plot(dataset, ENV, GEN, KW)
p2 = ge_plot(dataset, ENV, GEN, KW) + theme_grey() + coord_flip()
plot_grid(p1, p2)
```


# Joint regression analysis
@Eberhart1966 popularized the regression-based stability analysis. In these procedures, the adaptability and stability analysis is performed by means of adjustments of regression equations where the dependent variable is predicted as a function of an environmental index, according to the following model:

$$
\mathop Y\nolimits_{ij}  = {\beta _{0i}} + {\beta _{1i}}{I_j} + {\delta _{ij}} + {\bar \varepsilon _{ij}}
$$
where ${\beta _{0i}}$ is the grand mean of the genotype *i* (*i* = 1, 2, ..., I); ${\beta _{1i}}$ is the linear response (slope) of the genotype *i* to the environmental index; *Ij* is the environmental index (*j* = 1, 2, ..., *e*), where ${I_j} = [(y_{.j}/g)- (y_{..}/ge)]$, ${\delta _{ij}}$ is the deviation from the regression, and ${\bar \varepsilon _{ij}}$ is the experimental error.
The model is fitted with the function `ge_reg()`. The S3 methods `plot()` and `summary()` may be used to explore the fitted model.

```{r, fig.width=5, fig.height=4, warning = FALSE, message = FALSE}
reg_model = ge_reg(dataset, ENV, GEN, REP, KW)
plot(reg_model$KW)
summary(reg_model)
```


# Genotypic confidence index
@Annicchiarico1992 proposed that a stability method in which the stability parameter is measured by the superiority of the genotype in relation to the average of each environment, according to the following model:

$$
{Z_{ij}} = \frac{{{Y_{ij}}}}{{{{\bar Y}_{.j}}}} \times 100
$$
The genotypic confidence index of the genotype *i* ($W_i$) is then estimated as follows:

$$
W_i = Z_{i.}/e - \alpha \times sd(Z_{i.})
$$
Where $\alpha$ is the quantile of the standard normal distribution at a given probability error ($\alpha \approx 1.64$ at 0.05). The method is implemented using the function `Annicchiarico()`. The confidence index is estimated considering all environment, favorable environments (positive index) and unfavorable environments (negative index), as follows:

```{r, fig.width=5, fig.height=4, warning = FALSE, message = FALSE}
ann5 = Annicchiarico(dataset, ENV, GEN, REP, KW)
ann1 = Annicchiarico(dataset, ENV, GEN, REP, KW, prob = 0.01)
summary(ann5)
```


# Superiority index
The function `superiority()` implements the nonparametric method proposed by @Lin1988, which considers that a measure of cultivar general superiority for cultivar x location data is defined as the distance mean square between the cultivar's response and the maximum response averaged over all locations, according to the following model.

$$
P_i = \sum\limits_{j = 1}^n{(y_{ij} - y_{.j})^2/(2n)}
$$
where *n* is the number of environments

Similar then the genotypic confidence index, the superiority index is calculated by all environments, favorable, and unfavorable environments.

```{r, fig.width=5, fig.height=4, warning = FALSE, message = FALSE}
super = superiority(dataset, ENV, GEN, REP, KW)
summary(super)
```


# Environmental stratification
A method that combines stability analysis and environmental stratification using factor analysis was proposed by @murakami2004. This method is implemented with the function `ge_factanal()`, as follows:

```{r, fig.width=5, fig.height=4, warning = FALSE, message = FALSE}
fact = ge_factanal(dataset, ENV, GEN, REP, KW, mineval = 0.9)
plot(fact$KW)
summary(fact)
```


# References

