---
title: "Indexes for simultaneous selection"
author: "Tiago Olivoto"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
link-citations: true
bibliography: metanref.bib 
vignette: >
  %\VignetteIndexEntry{Indexes for simultaneous selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r global_options, include = FALSE}
knitr::opts_chunk$set(comment = "#", collapse = TRUE)
load("indexes.RData")


```


# Getting started
In this section we will use the data examples `data_ge` and `data_ge2`  provided in the **metan** package. For more information, please, see `?data_ge` and `?data_ge2`. Other data sets can be used provided that the following columns are in the dataset: environment, genotype, block/replicate and response variable(s). See the section [Rendering engine](#rendering) to know how HTML tables were generated.


```{r warning = FALSE, message = FALSE, echo=FALSE}
library(metan)
library(DT) # Used to make the tables
# Function to make HTML tables
print_table <- function(table, rownames = FALSE, digits = 3, ...){
  df <- datatable(table, rownames = rownames, extensions = 'Buttons',
                  options = list(scrollX = TRUE, 
                                 dom = '<<t>Bp>',
                                 buttons = c('copy', 'excel', 'pdf', 'print')), ...)
  num_cols <- c(as.numeric(which(sapply(table, class) == "numeric")))
  if(length(num_cols) > 0){
    formatSignif(df, columns = num_cols, digits = digits)
  } else{
    df
  }
}

```

# Multi-trait stability index
The function `mtsi()` is used to compute the multi-trait stability index (*MTSI*) proposed by @Olivoto2019a. The first argument is a model of the class `waasb` or `waas`. It is possible to compute the *MTSI* for both *WAASB* -stability only- and *WAASBY* -simultaneous selection for mean performance and stability.

## Based on stability only

In the following example, the selection of stable genotypes will consider five traits, namely, KW, NKE, PH, EH, and TKW. Note that the output of the function `waasb()` is passed to the function `mtsi()` by the forward-pipe operator `%>%`. Finally, the MTSI index is plotted using the function `plot()`.

```{r eval=FALSE}
library(metan)
MTSI_index <- data_ge2 %>%
              waasb(ENV, GEN, REP,
                    resp = c(KW, NKE, PH, EH, TKW)) %>%
              mtsi(verbose = FALSE, index = "waasb")

```


## Based on mean performance and stability

The following code considers that higher values for KW, NKE, TKW are better, and lower values for PH and EH are better. By using `wresp = 65`, the simultaneous selection for mean performance and stability will prioritize the mean performance (mean of the variables) rather than their stability across environments.

```{r eval=FALSE}
MTSI_index2 <- 
  data_ge2 %>%
  waasb(ENV, GEN, REP,
        resp = c(KW, NKE, PH, EH, TKW),
        mresp = c("h, h, l, l, h"),
        wresp = 65) %>% # Default is 50
  mtsi(SI = 20)

```


```{r, fig.height = 5, fig.width = 5.5, fig.align = "center", message=FALSE, warning=FALSE }
# Factor analysis
gmd(MTSI_index2, "FA") %>% print_table()
# Scores for genotypes
gmd(MTSI_index2, "scores_gen") %>% print_table()
# Selection differential for the 'waasby' index
gmd(MTSI_index2, "sel_dif_mps") %>% print_table()
# Selection differential for the waasb index
gmd(MTSI_index2, "sel_dif_stab") %>% print_table()
# Selection differential for the variables
gmd(MTSI_index2, "sel_dif_trait") %>% print_table()

plot(MTSI_index2)
```

# MGIDI index
The MGIDI index can be seen as the MTSI index with a weigth of 100 for mean performance. This index is computed with the function `mgidi()`. Here, we will use the example data `data_g()`. By default, all traits are assumed to be increased. To change this default, use the argument `ideotype`. For example, if for three traits, the first one is assumed to be decreased and the last two are assumed to be increased, use `ideotype = c("l, h, h")`.

```{r,   eval=FALSE}
mod <- gamem(data_g,
             gen = GEN,
             rep = REP,
             resp = everything())
mgidi_index <- mgidi(mod,
                     SI = 20) # Selection intensity

```

```{r fig.width=8, fig.height=4}
p1 <- plot(mgidi_index, SI = 20)
p2 <- plot(mgidi_index, type = "contribution")
arrange_ggplot(p1, p2)
```

# Smith-Hazel index
The Smith-Hazel index ([@Smith1936; @Hazel1943]) is computed with the function `Smith_Hazel()`. Users can compute the index either by declaring known genetic and phenotypic variance-covariance matrices or by using as inpute data, a model fitted with the function `gamem()`. In this case, the variance-covariance are extracted internally. The economic weights in the argument `weights` are set by default to 1 for all traits.

```{r  eval=FALSE}
smith <- Smith_Hazel(mod, SI = 20)

```




# FAI-BLUP index
The FAI-BLUP is a multi-trait index based on factor analysis and ideotype-design recently proposed by @Rocha2018. It is based on factor analysis, when the factorial scores of each ideotype are designed according to the desirable and undesirable factors. Then, a spatial probability is estimated based on genotype-ideotype distance, enabling genotype ranking [@Rocha2018]. Here we will use the mixed-model `mod` as inpute data. By default, the selection is made to increase the value of all traits. Change this default with the arguments `DI` and `UI`.


```{r, eval=FALSE}

fai <- fai_blup(mod, SI = 20)

```

```{r, fig.height = 5, fig.width = 5.5, fig.align = "center", message=F, warning=FALSE}
plot(fai)
```


# Comparing the indexes

The function `coincidence_indexes()` can be used to compute the coincidence index described by @Hamblin1986, among the multi-trait indexes implemented in `metan`. To do that, just use a comma-separated list of indexes as inpute data and inform the total number of genotypes.

```{r}
coincidence <- coincidence_index(mgidi_index, fai, smith, total = 13)
coincidence

```

We can also produce a Venn plot to show the relationships between the indexes

```{r}
MGIDI <- gmd(mgidi_index, "sel_gen")
FAI <- gmd(fai, "sel_gen")
SH <- gmd(smith, "sel_gen")

# Create the plot
venn_plot(MGIDI, FAI, SH, show_elements = TRUE)
```





_______________________________________________________________________________________________
# Rendering engine {#rendering}

This vignette was built with [pkgdown](https://pkgdown.r-lib.org/). All tables were produced with the package [`DT`](https://rstudio.github.io/DT/) using the following function.

```{r warning = FALSE, message = FALSE, eval=FALSE}

library(DT) # Used to make the tables
# Function to make HTML tables
print_table <- function(table, rownames = FALSE, digits = 3, ...){
  df <- datatable(table, rownames = rownames, extensions = 'Buttons',
                  options = list(scrollX = TRUE, 
                                 dom = '<<t>Bp>',
                                 buttons = c('copy', 'excel', 'pdf', 'print')), ...)
  num_cols <- c(as.numeric(which(sapply(table, class) == "numeric")))
  if(length(num_cols) > 0){
    formatSignif(df, columns = num_cols, digits = digits)
  } else{
    df
  }
}

```




# References

