---
title: "Indexes for simultaneous selection"
output: rmarkdown::html_vignette
link-citations: true
bibliography: metanref.bib 
vignette: >
  %\VignetteIndexEntry{Indexes for simultaneous selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r global_options, include = FALSE}
knitr::opts_chunk$set(comment = "", cache = TRUE)

```


# Getting started
In this section we will use the data examples `data_ge` and `data_ge2`  provided in the **metan** package. For more information, please, see `?data_ge` and `?data_ge2`. Other data sets can be used provided that the following columns are in the dataset: environment, genotype, block/replicate and response variable(s).

# FAI-BLUP index
The FAI-BLUP is a multi-trait index based on factor analysis and ideotype-design recentely proposed by @Rocha2018. It is based on factor analysis, when the factorial scores of each ideotype are designed according to the desirable and undesirable factors. Then, a spatial probability is estimated based on genotype-ideotype distance, enabling genotype ranking [@Rocha2018]. The BLUPs from genotypes are obtained with the function `waasb()` and then the function `fai_blup()` is used to compute the FAI-BLUP index.



```{r, fig.height = 5, fig.width = 5.5, fig.align = "center", message=F, warning=FALSE}
library(metan)
library(cowplot) # used to arrange the graphics
library(kableExtra) # Used to make the tables
library(dplyr)
# Function to make HTML tables
print_table = function(table){
  kable(table, "html", digits = 3) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                  font_size = 12)
}

FAI = data_ge2 %>%
      waasb(ENV, GEN, REP, c(KW, NKE, PH, EH)) %>%
      fai_blup(DI = c("max, max, max, min"),
               UI = c("min, min, min, max"),
               SI = 15)
plot(FAI)
    ```



# Multi-trait stability index
The function `mtsi()` is used to compute the multi-trait stability index (*MTSI*) proposed by @Olivoto2019a. The first argument is a model of the class `waasb` or `waas`. It is possible to compute the *MTSI* for both *WAASB* -stability only- and *WAASBY* -simultaneous selection for mean performance and stability.

## Based on stability only

In the following example, the selection of stable genotypes will consider five traits, namely, KW, NKE, PH, EH, and TKW. Note that the output of the function `waasb()` is passed to the function `mtsi()` by the forward-pipe operator `%>%`. Finally, the MTSI index is plotted using the function `plot()`.

```{r, fig.height = 5, fig.width = 5.5, fig.align = "center" , message=FALSE, warning=FALSE}
MTSI_index = data_ge2 %>%
             waasb(ENV, GEN, REP,
                   resp = c(KW, NKE, PH, EH, TKW)) %>%
             mtsi()
plot(MTSI_index)
    ```


## Based on mean performance and stability

The following code considers that higher values for KW, NKE, TKW are better, and lower values for PH and EH are better. By using `wresp = 65`, the simultaneous selection for mean performance and stability will prioritize the mean performance (mean of the variables) rather than their stability across environments.

```{r, fig.height = 5, fig.width = 5.5, fig.align = "center", message=FALSE, warning=FALSE }
MTSI_index2 = data_ge2 %>%
             waasb(ENV, GEN, REP,
                   resp = c(KW, NKE, PH, EH, TKW),
                   mresp = c(100, 100, 0, 0, 100),
                   wresp = rep(65, 5)) %>% # Default is 50
             mtsi(index = "waasby")
plot(MTSI_index2)
    ```

# References

