---
title: "Indexes for simultaneous selection"
output: rmarkdown::html_vignette
link-citations: true
vignette: >
  %\VignetteIndexEntry{Indexes for simultaneous selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



# Getting started
A dataset called `data_ge` is provided to make reproducible examples. For more information, please, see `?data_ge`. Other data sets can be used provided that the following columns are in the dataset: environment, genotype, block/replicate and response variable(s).

```{r warning = FALSE, message = FALSE}
# Importing data
library(METAAB)
dataset = data_ge
```


# FAI-BLUP index

The FAI-BLUP index [@Rocha2018] is an index for simultaneous selection based on multitrait. The BLUPs from genotypes are obtained with the function `WAASB()` and then the function `FAI.BLUP()` is used to compute the FAI-BLUP index.

```{r }
multivariate = WAASB(dataset,
                     resp = c(GY, HM),
                     gen = GEN,
                     env = ENV,
                     rep = REP)

FAI = FAI.BLUP(multivariate,
               SI = 10,
               DI = c("max", "max"), 
               UI = c("min", "min"))

plot(FAI)
    ```



