---
title: "Multi-environment Trial Analysis"
output: rmarkdown::html_vignette
fig_caption: yes
link-citations: true
bibliography: metanref.bib    
vignette: >
  %\VignetteIndexEntry{Multi-environment Trial Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Brief examples
The package [`kableExtra`](https://haozhu233.github.io/kableExtra/) and [`cowplot`](https://github.com/wilkelab/cowplot) were used to generate the tables and arranging the graphics of this vignette. The forward-pipe operators `%>%` and `%<>%` may be used in some functions, provided that the package [`magrittr`](https://magrittr.tidyverse.org/) is loaded. Brief examples will be run using the dataset `data_ge` that contains data on two variables assessed in 10 genotypes growing in 11 environments. For more details see `?data_ge`.

```{r, message=FALSE, warning=FALSE}
library(metan)
library(cowplot) # used to arrange the graphics
library(kableExtra) # Used to make the tables
library(magrittr) # for the forward-pipe operator %>%

# Function to make HTML tables
print_table = function(table){
  kable(table, "html", digits = 3) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 12)
}
str(data_ge)
```


# AMMI model
## Cross-validation procedures
The cross-validation procedures implemented in the \bold{metan} are based on the splitting of the original data into a training set and a validation set. The model is fitted using the training set and the predicted value is compared with the validation set. This process is iterated many times, say, 1000 times. The lesser the difference between predicted and validation data, the higher the predictive accuracy of the model. More information may be found [here.](https://tiagoolivoto.github.io/metan/articles/vignettes_CROSS-VALIDATION.html)

```{r, fig.height=3.5, fig.width=5}
CVAL = cv_ammif(data_ge,
                resp = GY,
                gen = GEN,
                env = ENV,
                rep = REP,
                nboot = 50,
                nrepval = 2)
plot(CVAL)
```


### Fitting the model
The AMMI model is fitted with the function `waas()`. Note that a model may be fitted easier and faster using `%>%` and assigning the arguments in the correct order (environment, genotype, replication and response(s) variable(s). The use of `%>%` make it the coding logical and clean especially when the output of a function is used as input in a second function. For more details, please see `?waas`.

```{r}
model <- data_ge %>% waas(ENV, GEN, REP, GY)
```

### Biplots

The well-known AMMI2 biplot may be obtained using the function `plot_scores()`. ggplot2-based graphics are obtained. Please, note that since `waas()` and `waasb()` functions allow analyzing multiple variables at the same time, e.g., `resp = c(v1, v2, ...)`, the output `model` is a list that in this case has one element, GY.

```{r, fig.height=8, fig.width=5,  message=FALSE, warning=FALSE}
p1 <- plot_scores(model$GY, axis.expand = 1.2)
p2 <- plot_scores(model$GY,
                  type = 1,
                  polygon = TRUE,
                  col.gen = "black",
                  col.env = "gray70",
                  col.segm.env = "gray70",
                  axis.expand = 1.5)
plot_grid(p1, p2,
          align = "v",
          labels = c("p1","p2"),
          ncol = 1)
```


### S3 method predict()
The S3 method `predict()` is implemented for objects of class `waas` and may be used to estimate the response of each genotype in each environment considering different number of Interaction Principal Component Axis (IPCA). As a example, we will use four IPCA (number of significant IPCAs) to estimate the GY using the previously fitted `model` object.

```{r }
predicted <- predict(model, naxis = 4)
print_table(head(predicted$GY))
```

# BLUP model
The implementation of linear-mixed effect models to predict the response variable in MET is made with `waasb()` function. The *mixed-effect version* of the already fitted AMMI model, where genotype and genotype-vs-environment interaction are assumed to have random effects is then obtained as follows.

```{r }
model2 <- data_ge %>% waasb(ENV, GEN, REP, GY)
```

The mixed-effect model fitted in `model2` has many outputs. For example, we can easely obtain the Likelihood Ration Test for random effects, the variance components, and the BLUPs for genotypes as follows:

## Residual plots
Several residual plots may be obtained applying the function `autoplot()` to an object of class `waasb` (or `waas`). 

```{r,fig.height=12, fig.width=4}
library(ggplot2)
autoplot(model2$GY, which = c(1, 2, 7), mfrow = c(3,1))
```

## Distributin of random effects
The distribution of the random effects may be obtained using the argument
`type = "re"`.

```{r,fig.height=8, fig.width=4}
autoplot(model2$GY, type = "re", which = c(1, 2), mfrow = c(2,1))
```



```{r }
print_table(model2$GY$LRT)

print_table(model2$GY$random)

print_table(model2$GY$ESTIMATES)

print_table(model2$GY$blupGEN)
```

An easy way to check the results in the R console is by using the function `summary()`


## Plotting the BLUPs for genotypes

```{r, fig.height=8, fig.width=4}
p1 <- plot_blup(model2$GY)
p2 <- plot_blup(model2$GY,
                prob = 0.1,
                col.shape  =  c("gray20", "gray80")) + coord_flip()
plot_grid(p1, p2,
          align = "v",
          labels = c("p1", "p2"),
          ncol = 1)
```


## BLUPS for genotype-vs-environment interaction

```{r }
print_table(head(model2$GY$BLUPgge))
```

## BLUP-based stability index
The function `Resende_indexes()` may be used to compute the harmonic mean of genotypic values (HMGV), the relative performance of the genotypic values (RPGV) and the harmonic mean of the relative performance of genotypic values (HMRPGV). See @Alves2018 for more details.

```{r }
index = model2 %>%
  Resende_indexes()
  print_table(index$GY)
```



# GGE model
## Fitting the model
The GGE model is fitted with the function `gge()`. This function produces a GGE model based on both a two-way table (in our case the object `table`) with genotypes in the rows and environments in columns, or a data.frame containing at least the columns for genotypes, environments and the response variable(s).

```{r echo = TRUE}
# Using a data.frame
gge_model = gge(data_ge,
                gen = GEN,
                env = ENV,
                resp = GY)

# Using a two-way table
ge_table = make_mat(data_ge, GEN, ENV, GY)
gge_model = gge(ge_table, table = TRUE)

```

## Visualizing the Biplot
The generic function `plot()` is used to generate a biplot using as input a fitted model of class `gge`.  The type of biplot is chosen by the argument `type` in the function.  Ten biplots type are available according to {@Yan2003}.

*`type = 1` A basic biplot.
*`type =  2` Mean performance vs. stability.
*`type =  3` Which-won-where.
*`type =  4` Discriminativeness vs. representativeness.
*`type =  5` Examine an environment.
*`type =  6` Ranking environments.
*`type =  7` Examine a genotype.
*`type =  8` Ranking gentoypes.
*`type =  9` Compare two genotypes.
*`type =  10` Relationship among environments.

```{r echo = TRUE, fig.height=3.5, fig.width=5, fig.align = "center", message=F, warning=F}
plot(gge_model)
```


# Parametric and non-parametric statistics
The metan package provides an easy way to compute several world-known stability methods.
Up to the version 1.0.3, the following stability methods are implemented:

* Joint regression analysis [@Eberhart1966], using the function `ge_reg()`
* Genotypic confidence index [@Annicchiarico1992], using the function `Annicchiarico()`
* Nonparametric superiority index [@Lin1988], using the function `superiority()`
* Ecovalence [@Wricke1965], using the function `ecovalence()`
* Stability analysis and environmental stratification [@murakami2004], using the function `ge_factanal()`

The easiest way to compute the above-mentioned stability indexes is by using the function `ge_stats()`. This function basically returns a summary of each method. If you are looking for more details from each method, like plot() and summary(), I suggest computing that methods using their own function.

# Computing the indexes

```{r }
stat_ge <- data_ge %>% ge_stats(ENV, GEN, REP, GY)
```
If you whant to export a summary of results, the simplest way is by using the function summary, as follows:


```{r, eval=FALSE }
summary(stat_ge, export = TRUE)
```
This comand will create a text file called *ge_stats summary.txt* in the current working directory.


# Extending the metan package

* The complete functionality of the metan package, combining theory, programming, and examples with outputs is described [here](https://tiagoolivoto.github.io/metan/index.html). Suggestions and criticisms to improve the quality and usability of the package are welcome!

* Feel free to report a bug at https://github.com/TiagoOlivoto/metan/issues

